---
title: "Lecture Week 13: GLM"
format: 
  html:
    embed-resources: true
editor: visual
toc: true
toc-depth: 3
number-sections: true
theme:
  light: flatly
  dark: darkly
execute: 
  cache: true
---

```{r}
#| label: setup
#| include: false

# libraries
library(tidyverse)
library(Sleuth3)

# data objects
elephant <- case2201
```

# Overview

-   Generalized linear models are useful when your response variable is not normally distributed.

-   GLMs are often an alternative to transformations of data

-   GLMs work by using linear predictors, a variance function, and a link function

-   Linear predictors are what we have used before - they can be categorical or continuous

-   The linear models we have used so far use the normal distribution for our variance function (hence meeting the assumptions previously discussed), but now we can pick different functions

    -   Here we will look at the $Poisson(\lambda)$ and the $Binomial(n, p)$ distributions

-   The link function is new - this takes the place of transformations, but instead of transforming the raw data (as before), it transforms the predictions made by the linear component

# Setup

### Libraries

```{r}
#| eval: false
# libraries
library(tidyverse)
library(Sleuth3)

```

### Data

```{r}
#| eval: false
# data objects
elephant <- case2201
```

# Lecture Outline

## Poisson distribution

-   The Poisson distribution is useful when the variable takes on discrete, positive, whole numbers

-   This is often the count of something

    -   Minimum bound is 0

    -   Maximum bound is $+\infty$

-   The Poisson is described by a single parameter $\lambda$

-   Therefore, we would say that:

$$
Y \sim Poisson(\lambda)
$$

-   Recall that a normal distribution is defined with two parameters: $N(\mu, \sigma)$, where $\mu$ is the average or expected value, and $\sigma$ is the spread of the data, but it is a fixed entity

-   With a poisson, $\lambda$ described the expected value and the variance or noise around that value:

-   

$$
\lambda = E(Y) = Var(Y) 
$$

-   So as the expected value increases, so does the variation

```{r}
#| message: false
#| warning: false
pois_df <- expand_grid(iter = 0:30,
                       lambda = c(1, 5, 10, 20)) |>
  mutate(y = dpois(iter, lambda))

ggplot(pois_df,
       aes(y = y, 
           x = iter, 
           fill = as_factor(lambda))) +
  geom_density(stat = "identity",
               alpha = 0.5) +
  scale_fill_viridis_d(option = "turbo") +
  theme_bw()

ggplot(pois_df,
       aes(y = y, 
           x = iter, 
           fill = as_factor(lambda))) +
  geom_histogram(stat = "identity",
               alpha = 0.35,
               position = "identity",
               color = "black") +
  scale_fill_viridis_d() +
  theme_minimal()
```

## Elephant Data

-   `case2201` is observational data on the number of mates that male elephants have and their age

-   Male elephants are sexually mature at \~15 years old, but can rarely out compete older, larger males for access to females

```{r}
as_tibble(elephant)
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() 
```

-   This looks like a positive relationship, and we can put a linear regression line here:

```{r}
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() +
  geom_smooth(method = "lm")
```

-   Note that if we extended the line, it would quickly predict "negative" mates, which is not possible

```{r}
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() +
  geom_smooth(method = "lm",
              fullrange = TRUE) +
  scale_x_continuous(limits = c(0, 50))+
  scale_y_continuous(limits = c(-10, 10))
```

## Poisson Regression with `glm()` 

-   The `lm()` function produces linear models

-   The `glm()` function produces generalized linear models

-   The syntax is generally the same with a formula and a data argument

-   What is added is a `family =` argument which allows you to designate other distribution families.

-   We will set `family = poisson`

    -   There is also a `link =` argument inside of the family call

    -   The default link for Poisson is `log`

    -   Other links exist, but we will just stick with the defaults here

    -   Note that the back-transformation for a `log()` value is to exponentiate the value; `exp()` in R

        -   i.e., `log(4)` = 1.386294

        -   `exp(1.386296)` = 3.999999 = 4

```{r}
ele_glm <- glm(Matings ~ Age,
               data = elephant,
               family = poisson(link = "log"))
summary(ele_glm) 
```

-   Before we go into the model estimates, we first need to ask if Poisson is a reasonable fit for this data.

-   A Poisson distributed variable should have its mean and variance be approximately equal.

-   We can check this by looking at the ratio of residual deviance to the degrees of freedom

    -   A perfect ratio would be 1

    -   General rule of thumb is anything \> 2 could be problematic

```{r}
51/39
```

-   This model looks good

## Describe the link function here?

## Working with the model

```{r}
summary(ele_glm)
```

-   We have coefficient estimates for our intercept and $\beta_1$ So we can write out our equation

-   An important note though, is that the equation will predict the number of mates on the *transformed* scale

    -   With every 1-year increase in age, a male will have an increase of 0.069 mates *on the log-scale*

    -   `exp(0.069)` = 1.07. This is **not** an increase of 1.07 mates. It's an increase of 7% more mates than the year before

    -   This is a percentage change, not an absolute change

-   It is usually easier to estimate the response on the log-scale, and then back-transform our result to get back on the same scale as the original data

    -   i.e., the number of mates

$$
Log(Mates) = \beta_0 + \beta_1 * Age
$$

-   Let's fill out the equation, and estimate the number of mates for a 40 year old male

$$
Log(Mates) = -1.58 + 0.069 * 40
$$

```{r}
# log-scale
-1.58 + 0.069*40
```

-   To put this on the "real" scale, we need to exponentiate

$$
e^{1.18}
$$

```{r}
exp(1.18)
# "real" scale
```

-   A 40 year old male elephant will have had an average of **3.25 mates**

::: callout-note
## Prediction with GLM's

-   It is possible to use `predict()` with `newdata` for GLM models

-   There are a few extra steps to accurately estimate the interval (either confidence or prediction)

-   I'm not going to go into the details here, but there are a number of resources online which explain it in detail

-   For the home work problems in this class, I will only ask you to calculate the estimated response for a given value of a predictor variable one at a time

    -   i.e., like we did above

    -   Plug in one x-value to the equation

    -   calculate the estimate on the log-scale

    -   Exponentiate to get the estimate back on the same scale as the data
:::

## Plot model results

-   We can plot the model results with ggplot

```{r}
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() +
  geom_smooth(method = "glm",
              method.args = list("poisson")) 
```

-   Note the "curved" nature of the line

-   And the increasing uncertainty at higher x-values

    -   Reflecting the increasing variance of a Poisson distribution

-   And when we extend our model, the fitted line approaches 0 but never goes negative.

```{r}
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() +
  geom_smooth(method = "glm",
              method.args = list("poisson"),
              fullrange = TRUE) +
  scale_x_continuous(limits = c(0, 50))+
  scale_y_continuous(limits = c(-10, 10))
```

# Summary
