---
title: "Lecture Week 13: GLM"
format: 
  html:
    embed-resources: true
editor: visual
toc: true
toc-depth: 3
number-sections: true
theme:
  light: flatly
  dark: darkly
execute: 
  cache: true
---

```{r}
#| label: setup
#| include: false

# libraries
library(tidyverse)
library(Sleuth3)

# data objects
elephant <- case2201
beetle <- read.csv(here::here("data/beetle.csv"))
```

# Overview

-   Generalized linear models are useful when your response variable is not normally distributed.

-   GLMs are often an alternative to transformations of data

-   GLMs work by using linear predictors, a variance function, and a link function

-   Linear predictors are what we have used before - they can be categorical or continuous

-   The linear models we have used so far use the normal distribution for our variance function (hence meeting the assumptions previously discussed), but now we can pick different functions

    -   Here we will look at the $Poisson(\lambda)$ and the $Binomial(n, p)$ distributions

-   The link function is new - this takes the place of transformations, but instead of transforming the raw data (as before), it transforms the predictions made by the linear component

# Setup

### Libraries

```{r}
#| eval: false
# libraries
library(tidyverse)
library(Sleuth3)

```

### Data

```{r}
#| eval: false
# data objects
elephant <- case2201
beetle <- read.csv(here::here("data/beetle.csv"))
```

# Lecture Outline

# Poisson distribution

-   The Poisson distribution is useful when the variable takes on discrete, positive, whole numbers

-   This is often the count of something

    -   Minimum bound is 0

    -   Maximum bound is $+\infty$

-   The Poisson is described by a single parameter $\lambda$

-   Therefore, we would say that:

$$
Y \sim Poisson(\lambda)
$$

-   Recall that a normal distribution is defined with two parameters: $N(\mu, \sigma)$, where $\mu$ is the average or expected value, and $\sigma$ is the spread of the data, but it is a fixed entity

-   With a poisson, $\lambda$ described the expected value and the variance or noise around that value:

$$
\lambda = E(Y) = Var(Y) 
$$

-   So as the expected value increases, so does the variation

```{r}
#| message: false
#| warning: false
pois_df <- expand_grid(iter = 0:30,
                       lambda = c(1, 5, 10, 20)) |>
  mutate(y = dpois(iter, lambda))

ggplot(pois_df,
       aes(y = y, 
           x = iter, 
           fill = as_factor(lambda))) +
  geom_density(stat = "identity",
               alpha = 0.5) +
  scale_fill_viridis_d(option = "turbo") +
  theme_bw()

ggplot(pois_df,
       aes(y = y, 
           x = iter, 
           fill = as_factor(lambda))) +
  geom_histogram(stat = "identity",
               alpha = 0.35,
               position = "identity",
               color = "black") +
  scale_fill_viridis_d() +
  theme_minimal()
```

## Elephant Data

-   `case2201` is observational data on the number of mates that male elephants have and their age

-   Male elephants are sexually mature at \~15 years old, but can rarely out compete older, larger males for access to females

```{r}
as_tibble(elephant)
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() 
```

-   This looks like a positive relationship, and we can put a linear regression line here:

```{r}
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() +
  geom_smooth(method = "lm")
```

-   Note that if we extended the line, it would quickly predict "negative" mates, which is not possible

```{r}
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() +
  geom_smooth(method = "lm",
              fullrange = TRUE) +
  scale_x_continuous(limits = c(0, 50))+
  scale_y_continuous(limits = c(-10, 10))
```

## Poisson Regression with `glm()`

-   The `lm()` function produces linear models

-   The `glm()` function produces generalized linear models

-   The syntax is generally the same with a formula and a data argument

-   What is added is a `family =` argument which allows you to designate other distribution families.

-   We will set `family = poisson`

    -   There is also a `link =` argument inside of the family call

    -   The default link for Poisson is `log`

    -   Other links exist, but we will just stick with the defaults here

    -   Note that the back-transformation for a `log()` value is to exponentiate the value; `exp()` in R

        -   i.e., `log(4)` = 1.386294

        -   `exp(1.386296)` = 3.999999 = 4

```{r}
ele_glm <- glm(Matings ~ Age,
               data = elephant,
               family = poisson(link = "log"))
summary(ele_glm) 
```

-   Before we go into the model estimates, we first need to ask if Poisson is a reasonable fit for this data.

-   A Poisson distributed variable should have its mean and variance be approximately equal.

-   We can check this by looking at the ratio of residual deviance to the degrees of freedom

    -   A perfect ratio would be 1

    -   General rule of thumb is anything \> 2 could be problematic

```{r}
51/39
```

-   This model looks good

## Describe the link function here?

## Working with the model

```{r}
summary(ele_glm)
```

-   We have coefficient estimates for our intercept and $\beta_1$ So we can write out our equation

-   An important note though, is that the equation will predict the number of mates on the *transformed* scale

    -   With every 1-year increase in age, a male will have an increase of 0.069 mates *on the log-scale*

    -   `exp(0.069)` = 1.07. This is **not** an increase of 1.07 mates. It's an increase of 7% more mates than the year before

    -   This is a percentage change, not an absolute change

-   It is usually easier to estimate the response on the log-scale, and then back-transform our result to get back on the same scale as the original data

    -   i.e., the number of mates

$$
Log(Mates) = \beta_0 + \beta_1 * Age
$$

-   Let's fill out the equation, and estimate the number of mates for a 40 year old male

$$
Log(Mates) = -1.58 + 0.069 * 40
$$

```{r}
# log-scale
-1.58 + 0.069*40
```

-   To put this on the "real" scale, we need to exponentiate

$$
e^{1.18}
$$

```{r}
exp(1.18)
# "real" scale
```

-   A 40 year old male elephant will have had an average of **3.25 mates**

## Use `predict()`; designate output

-   `predict()` allows us to designate if we want the estimate on the `"link"` or the `"response"` scale

```{r}
predict(ele_glm,
        newdata = list(Age = 40),
        type = "link")
predict(ele_glm,
        newdata = list(Age = 40),
        type = "response")
```

-   These values reasonably match our manual calculations above (with rounding)

::: callout-note
## GLM Intervals with `predict()`

-   Note that it is not possible to use get intervals out of `predict()` when using GLM models

-   You can calculate them manually, and there also appear to be other packages which you can download which will do it.

-   I'm not going to go into the details here, but there are a number of resources online which explain it in detail

-   For the home work problems in this class, I will only ask you to calculate the estimated response for a given value of a predictor variable, and not any of the intervals
:::

## Plot model results

-   We can plot the model results with ggplot

```{r}
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() +
  geom_smooth(method = "glm",
              method.args = list("poisson")) 
```

-   Note the "curved" nature of the line

-   And the increasing uncertainty at higher x-values

    -   Reflecting the increasing variance of a Poisson distribution

-   And when we extend our model, the fitted line approaches 0 but never goes negative.

```{r}
ggplot(elephant,
       aes(x = Age, 
           y = Matings)) +
  geom_point() +
  geom_smooth(method = "glm",
              method.args = list("poisson"),
              fullrange = TRUE) +
  scale_x_continuous(limits = c(0, 50))+
  scale_y_continuous(limits = c(-10, 10))
```

# Binomial Distribution

-   The binomial distribution is useful when the response variable is bounded by 0 and 1.

-   Proportion data commonly takes this form

-   Proportions can be used themselves (i.e., 0.25, 0.5, etc.)

-   However, when the number of trial varies, it is best to supply a "weight" column

    -   For example, 1/2 and 50/100 are the same proportion, but the 50/100 has more weight, or is more consistent

        -   It would be hard to have 50/100 by random chance, but 1/2 by random chance may not be unexpected

-   This way, the model can weight data with more trials more heavily

-   The binomial distribution is described as:

$$
Y \sim Binomial(n, p)
$$

Where $n$ is the number of trials and $p$ is the probability of success

-   Note that you can define "success" either way, but it'll change your interpretation.

    -   Be sure to define it at the beginning and then stay consistent

-   The expected value of a variable with $Binomial(n, p)$ is $E[Y] = np$

-   Variance is: $Var(Y) = np(1-p)$

```{r}
#| warning: false
binom_df <- expand_grid(success = 0:10,
                       size = c(10),
                       p = c(0.1, 0.5, 0.9)) |>
  mutate(y = dbinom(success, size, p))

ggplot(binom_df,
       aes(y = y, 
           x = success)) +
  geom_histogram(stat = "identity",
               alpha = 0.75,
               color = "black") +
  theme_bw() +
  facet_wrap(~p, labeller = labeller(p = label_both))


```

-   Note that the values can "pile up" on 0 or 1 when $p$ is low or high, respectively.

## Beetle data  

-   The `beetle.csv` contains examining the lethal effects of a pesticide on the flour beetle *Tribolium confusum*. The data was originally published in 1935 and has been used extensively in introductory statistics texts

```{r}
as_tibble(beetle)
```

-   Let's plot the data and see what a binomial distribution can look like

```{r}
ggplot(beetle, 
       aes(x = Dose, 
           y = Mortality_rate)) +
  geom_point()
```

-   you can see that this looks linear, but note the slight curves near 0 and 1

-   If we were to put a line through this, we would find negative mortality rates at doses \~\<50, and mortality rates \>100% at doses \~\>80

    -   Both of these are impossible

-   You might think that we could just use a linear regression and only interpolate to avoid these issues

-   Let's jump ahead a bit and show the Binomial regression line to emphasize the curved relationship and compare it to a normal regression line

```{r}
ggplot(beetle, 
       aes(x = Dose, 
           y = Mortality_rate)) +
  geom_point(size = 3) +
  geom_smooth(method = "glm",
              method.args = list(binomial),
              aes(weight = Number_tested)) +
  geom_smooth(method = "lm",
              color = "red",
              fill = "pink") +
  theme_bw()
```

## Fit a glm model

-   We can use `glm()` again, but this time we will designate `family = binomial`

-   We will use the default link function of `logit` for a binomial regression

    -   As before, there are other links, but we won't explore them here

```{r}
beetle_glm <- glm(Mortality_rate ~ Dose,
                  data = beetle, 
                  family = binomial(link = "logit"), 
                  weight = Number_tested)
beetle_glm
```

### The logit link function

$$ 
Logit(p) = log(\frac{p}{1-p}) 
$$

### Logits, log-odds, probability

-   The `logit` link function puts the prediction on the logit scale

-   Logits are weird

-   Logits are the natural logarithm of the odds, and the odds are a ratio of the successes to failures

-   Let's say we had 10 trials, which resulted in 5 successes and 5 failures

-   So the odds are 5/5, and the logit is the log of that:

```{r}
log(5/5)
```

-   One important logit to remember is 0, which equates to a 50% probability

    -   Negative logits means lower odds

    -   positive logits means higher odds

| Probablity | Odds (success / failure) | Logits (log-odds) |
|------------|--------------------------|-------------------|
| 0.01       | 0.010101                 | -4.595            |
| 0.2        | 0.2500                   | -1.386            |
| 0.5        | 1.00000                  | 0.000             |
| 0.8        | 4.0000                   | 1.386             |
| 0.99       | 99.0000                  | 4.595             |

: Table comparing probability, odds, and log odds (logits)

-   If you `exp()` your coefficients, that will give you the odds.

## Working with the model

-   our model coefficients are:

```{r}
coef(beetle_glm)
```

-   The $\beta_1$ coefficient is \> 0.0, indicating that probability of success (i.e., death) increases as dose increases, which should make sense.

<!-- -->

-   We can estimate the response on the logit-scale when `Dose = 60` by plugging in the numbers

$$
\eta = Logit(Mortality) = -14.578 + 0.2455 * 60 
$$

```{r}
eta <- coef(beetle_glm)[1] + coef(beetle_glm)[2] * 60
eta
```

-   To back transform to the original scale, we need to do the following:

$$
Mortality = \frac{exp(\eta)}{1 + exp(\eta)}
$$

```{r}
exp(eta) / (1+exp(eta))
```

## Use `predict()`; designate output

-   We can simplify this by using predict

-   `predict()` allows us to designate if we want the estimate on the `"link"` or the `"response"` scale

```{r}
predict(beetle_glm,
        newdata = list(Dose = 60),
        type = "link")
predict(beetle_glm,
        newdata = list(Dose = 60),
        type = "response")
```

-   We can add multiple predictions by creating a new data frame

-   Predict the response on the link scale for doses of 50, 60, and 70

```{r}
predict(beetle_glm,
        newdata = list(Dose = c(50, 60, 70)),
        type = "response")
```

::: callout-note
## Homework has two predictor variables

-   When you predict the response in the binomial homework problem, there are 2 predictor variables.

-   You have two options

    1.  Make a dataframe which `expand.grid()` like we did in the ANCOVA notes

    2.  Add another variable in the `list()` argument in `predict()`

        -   If you go this route, you need to make sure the lengths are the same

        -   i.e., `Dose = c(1, 2, 3)` and `other_x = c("a", "b", "c")`
:::

# Summary

-   GLM's are useful for when our response variable is not normal

-   They can be compared to transforming the data, but are generally better since they account for differences in distributions and variance structures.

-   We looked at two distributions here:

    -   Poisson - Positive whole numbers

    -   Binomial - continuous numbers bounded by 0 and 1

-   There are many other distribution families that are available for use in `glm()`

-   Likewise, we looked at the default link functions, but more are available, making glm's a flexible and powerful analytical tool.
