---
title: "Lecture Week Number: Topic"
html:
    embed-resources: true
    link-external-newwindow: true
editor: visual
toc: true
toc-depth: 2
number-sections: true
theme:
  light: flatly
  dark: darkly
---

```{r}
#| label: setup
#| include: false

# libraries
library(tidyverse)

# data objects
library(tidyverse)
bt_samp <- read_csv(here::here("data/body_temp_hr.csv"))
caterpillar <- read_csv(here::here("data/caterpillar.csv"))
```

# Overview

Introduce T-tests including their history and application. We will discuss the concept of Null hypothesis testing. Finally, we will practice applying three main types of T-tests. 


# Setup

### Libraries

```{r}
#| eval: false
library(tidyverse)
```

### Data

```{r}
#| eval: false
bt_samp <- read_csv("data/body_temp_hr.csv")
caterpillar <- read_csv("data/caterpillar.csv")
```

# Lecture Outline

## T-tests
### T-test overview

- Is my sample different from...:  

  + A Value of interest?  
  
  + Another sample of data?   
  
- General formula of t-test:

$$t_{statistic} = \frac{difference}{variation}$$

  + No difference: $t_{statistic} = 0$  
  
  + There is a difference: $t_{statistic} \ne 0$  


### History
- t-test developed by William Sealy Gossett  

- Worked at Guinness, wanted to test quality of stout  

- Very small sample sizes ( $n \le 4$ )  

- Uses the t-distribution, which can be thought of as a small-sample-size version of a standard normal distribution  

- Published under pseudonym *Student* because company policy prevented employees from publishing scientific papers  


### T-distribution
- t-distribution is a small-sample-size version of standard normal distribution  

- with fewer samples, there is more uncertainty  

- this uncertainty is represented in "fat tails"  
```{r t, fig.width=8, fig.height=5.5}
curve(dt(x, 1), -5, 5, ylim=c(0, 0.5),
      xlab="t value", ylab="Probability density",
      cex.lab=1.5, lwd=1.5)
curve(dt(x, 5), -5, 5, add=TRUE, col="blue", lwd=1.5)
curve(dnorm(x, 0, 1), -5, 5, add=TRUE,  lty=3, lwd=1.5)
legend(-5, 0.5, c("Standard normal distribution", "t distribution (df=5)", "t distribution (df=1)"),
       lty=c(3, 1, 1), col=c("black", "blue", "black"), lwd=1.5)
```


### Types of T-tests
- One-sample
  - Is my sample different from a value of interest?
  - i.e., stream pH, human body temperature
- Two-sample
  - Are my two samples of data different?
  - i.e., pH in mined vs. un-mined streams? IS body temperature different between males and females, or sick and healthy groups?
- Paired
  - Are two measurements on the same subject different?
  - i.e., before/after drug treatment, or two streams (mined/un-mined) in one valley

##  Null Hypothesis Testing (NHT)
> Formal approach to assessing if sample data represents a statistically significant difference, or is due to sampling error.  

- requires two hypotheses: 

  - The null, $H_0$, which is no difference/relationship/effect  
  
  - Alternative, $H_A$, which is that there is some difference/relationship/effect  
  
  - Hypotheses refer to the **populations**  


##  One-sample T-Test  

### Hypotheses
### General Procedure
### Manual Procedure
### `lm()` Procedure
### Notes on P-values
### Interpretation

##  Two-sample T-Test
### `lm()` Procedure
-   Predictor needs to be a char. | fact.
    -   HW
### NHT for $\beta_0$ and $\beta_1$ 
### interpretation


##  Paired T-Test
### Hypotheses
### Data setup
-   Subtract columns
-   Lag calculation using `diff()`
### `lm()` Procedure
### Interpretation

# Summary
