---
title: "HW 10 Quarto"
execute: 
  echo: false
  eval: false
  message: false
  warning: false
format: html
editor: visual
toc: true
toc-depth: 2
embed-resources: true
theme:
  light: flatly
  dark: darkly
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
finch <- read.csv(here::here("data/galapago-finches.csv"))
```

# Learning Objectives

-   You will make your first Quarto document and render the final product as an html document

-   Practice editing the YAML header

-   Change text formats (headings, fonts, lists)

-   Write mathematical equations

-   Produce simple summaries and plots of `finch` data

-   Run a simple statistical analysis and interpret the results

# Setup

-   Start a new quarto document

-   **Title**: HW 10 Quarto

-   **Author**: Put your name in the author section

# Problems

## Problem 1: preliminaries

Edit the YAML header in your R Markdown file to include the following:

1.  Add the code to include a table of contents with `toc_depth : 2`\

2.  Add a level 1 Header called "My level 1 header"\

3.  Add a level 2 Header called "My level 2 header"\

4.  Add a level 3 Header called "My level 3 header"\

5.  click `Render` to ensure that an html document is created.

    -   Double check that the author field is correct

    -   Make sure that you have three headers, one of each level (1-3)

    -   Make sure that the table of contents at the top of your document *only* includes the first two levels.

-   **NOTE** You may want to `render` your document after you complete each problem to ensure that the out put format is correct.

## Problem 2: Lists

1.  Create a level 1 header titled "Problem 2: Lists".

2.  Create a level 2 header titled "My favorite animals". Then create a numbered list, ranking your top three favorite animals.

3.  Create a level 2 header called "Why use Quarto?". Then make a bulleted list (min 3 bullets) with a few of the benefits for using Quarto in your own words.

4.  Create a level 2 header called "Sub bulleted list". Then create a bulleted list (min 2 main bullets) with some of your favorite hobbies. For each main bullet, add at least one sub bullet briefly explaining why you like that hobby.

## Problem 3: Equations

1.  Create a level 1 header titled "Problem 3: Equations".
2.  Create an inline equation which produces the following: $x \sim Normal(\mu = 0, \sigma = 1)$
    -   **hint** the $\sim$ symbol can be added by typing `\sim` in your equation
3.  Create a block (AKA display) equation for a simple linear model. Your equation should look like this:
    1.  **Hint** the greek letters in use are beta and epsilon

$$ Y = \beta_0 + \beta_1 * X + $$

4.  Create a block equation that looks like this:

$$
HA: \mu_{reference} \neq \mu_{treatment}
$$

5.  Create the following block equation:
    -   **Hint** the symbol on the top is capital Sigma
    -   You can add a bar above a variable with `\bar{y}`

$$
s^2 = \frac{\Sigma(y_i - \bar{y})^2}{n-1}
$$

## Problem 4: Code Chunks

1.  Add a code chunk which loads the tidyverse and reads in the galapago-finches.csv file as an object called `finch` . At the end of the code chunk, display the `finch` object as a tibble

```{r}
library(tidyverse)
finch <- read.csv(here::here("data/galapago-finches.csv"))
as_tibble(finch)
```

2.  Add a code chunk which adds a new column. Name the column `year_factor` and create it by converting the original `year` column into a factor. Make sure to save this calculation as a new object called `finch2` , and display it as a tibble

```{r}
finch2 <- finch |>
  mutate(year_factor = factor(year))
as_tibble(finch2)
```

3.  Using `finch2`, make a scatterplot with `bdpeth` on the x-axis and `blength` on the y-axis. Also map the `shape` of the points to change with `year_factor`, and the `color` of the points to change with `species`. Also, add a line of best fit to the plot by using `stat_smooth(method = "lm")`

```{r}
ggplot(finch2,
       aes(x = bdepth,
           y = blength,
           color = species, 
           shape = year_factor)) +
  geom_point() +
  stat_smooth(method = "lm")
```

## Problem 5: Miscellany

1.  Create a level 1 header titled "Problem 5: Miscellany".
2.  Make a line with the following: "My CMU email address is:". Then, include your email address as a hyperlink.
3.  Include a line of text which includes at least one **bold** and one *italicized* word as well as one subscript and one superscript.

## Submission

Once your homework file is completed, you will need to render the document into an html. **Be sure to go over the rendered document to see if all the answer formats are correct.**

For credit, you will submit both the .qmd and .html files on D2L.
