---
title: "HW 10 Quarto"
author: "Justin Pomeranz"
execute: 
  echo: true
  eval: true
  message: false
  warning: false
format: html
editor: visual
toc: true
toc-depth: 2
embed-resources: true
theme:
  light: flatly
  dark: darkly
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
finch <- read.csv(here::here("data/galapago-finches.csv"))
```

# Problems

## Problem 1: preliminaries

# My level 1 header

## My level 2 header

### My level 3 header

## Problem 2: Lists

## My favorite animals

1.  Bugs (obviously)
2.  Here's another bug
3.  And a third

## Why Use Quarto

-   You can make lots of outputs

-   You can incorporate code/analysis/results directly in (don't need to copy/paste plots and stuff every time analysis changes

-   Focus on writing, not formatting

## Sub Bulleted list

-   Mountain Running

    -   Get to see all the sites on a backpacking trip without having to carry so much stuff

    -   Makes me feel cool

-   Playing the banjo

    -   Also makes me feel cool

## Problem 3: Equations

inline: $x \sim Normal(\mu = 0, \sigma = 1)$

Block equations:

$$ Y = \beta_0 + \beta_1 * X + $$

$$
HA: \mu_{reference} \neq \mu_{treatment}
$$

$$
s^2 = \frac{\Sigma(y_i - \bar{y})^2}{n-1}
$$

## Problem 4: Code Chunks

1.  Add a code chunk which loads the tidyverse and reads in the galapago-finches.csv file as an object called `finch` . At the end of the code chunk, display the `finch` object as a tibble

```{r}
library(tidyverse)
finch <- read.csv(here::here("data/galapago-finches.csv"))
as_tibble(finch)
```

2.  Add a code chunk which adds a new column. Name the column `year_factor` and create it by converting the original `year` column into a factor. Make sure to save this calculation as a new object called `finch2` , and display it as a tibble

```{r}
finch2 <- finch |>
  mutate(year_factor = factor(year))
as_tibble(finch2)
```

3.  Using `finch2`, make a scatterplot with `bdpeth` on the x-axis and `blength` on the y-axis. Also map the `shape` of the points to change with `year_factor`, and the `color` of the points to change with `species`. Also, add a line of best fit to the plot by using `stat_smooth(method = "lm")`

```{r}
ggplot(finch2,
       aes(x = bdepth,
           y = blength,
           color = species, 
           shape = year_factor)) +
  geom_point() +
  stat_smooth(method = "lm")
```

## Problem 5: Miscellany

1.  
2.  my cmu email address is [jpomeranz\@colroadomesa.edu](jpomeranz@colroadomesa.edu)
3.  Include a line of text which includes at least one
    1.  **bold**
    2.  *italicized*
    3.  subscript H~2~O
    4.  superscript M^2^

## 
