---
title: "Key_template"
execute: 
  echo: true
  eval: true
  message: false
  warning: false
format: html
editor: visual
toc: true
toc-depth: 2
embed-resources: true
number-sections: true
---

::: callout-warning
## Instructions for writing KEY and HW

1.  update libraries and data
    1.  In setup block (hidden) and in **SETUP** sections below
2.  Once Key is complete
    1.  Delete this instruction callout
    2.  Render
3.  After key is complete
    1.  "save as" appropriate HW in `assignment/` folder
    2.  Change `echo:` and `eval:` to `false` in YAML
    3.  Render assignment
        1.  Ensure that code and output are not visible
:::

```{r}
#| label: setup
#| include: false

library(tidyverse)

finch <- read.csv(here::here("data/galapago-finches.csv"),
                  stringsAsFactors = TRUE)
```

# Learning Objectives

# Setup

## Libraries

```{r}
#| eval: false
#| echo: true

library(tidyverse)
```

## Data

-   Download the "galapago-finches.csv" from D2L

-   Place it in your `data/` folder

-   Load the data into R with the following:

```{r}
#| eval: false
#| echo: true

finch <- read.csv("data/galapago-finches.csv",
                  stringsAsFactors = TRUE)

```

# Problems
