---
title: "Key_template"
execute: 
  echo: true
  eval: true
  message: false
  warning: false
format: html
editor: visual
toc: true
toc-depth: 2
embed-resources: true
theme:
  light: flatly
  dark: darkly
---

::: callout-warning
## Instructions for writing KEY and HW

1.  update libraries and data
    1.  In setup block (hidden) and in **SETUP** sections below
2.  Add graded comment to code block and as a 2nd-level header above problem
3.  Once Key is complete
    1.  Delete this instruction callout
    2.  Render
4.  After key is complete
    1.  "save as" appropriate HW in `assignment/` folder
    2.  Change `echo:` and `eval:` to `false` in YAML
    3.  Delete GRADED headers
    4.  Render assignment
        1.  Ensure that code and output are not visible
:::

```{r}
#| label: setup
#| include: false
#| eval: true

library(tidyverse)

finch <- read.csv(here::here("data/galapago-finches.csv"),
                  stringsAsFactors = TRUE)
```

# Overview

For your final assignment, you will need to find a data set (or data sets) from a published paper to perform statistical analyses on.

# Setup

## Libraries

-   Load all libraries that you need

-   At a minimum, you will want to use the `tidyverse`

-   Look through old lecture notes and load any other appropriate libraries

```{r}
#| eval: false
#| echo: true

library(tidyverse)
```

# Data

-   The first thing you will need to do is to locate a data set or multiple data sets to analyze.

-   Once you find data, you will save it as a .csv in your `"data/"` folder and load it appropriately in your Quarto document.

## Finding Data

-   [DataDRYAD](https://datadryad.org/)

# Problems
